---
title: Hierarchical Time&nbsp;Series Forecasting in Emergency Medical Services
author: Rob J Hyndman\newline Bahman Rostami-Tabar
pdf-engine: pdflatex
fig-width: 9
fig-height: 4.5
format:
  beamer:
    theme: monash
    aspectratio: 169
    fontsize: 14pt
    section-titles: false
    knitr:
      opts_chunk:
        dev: "CairoPDF"
include-in-header: header.tex
keep-tex: true
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| cache: false
source(here::here("setup.R"))
```

## Outline

\vspace*{0.4cm}\tableofcontents

# Data

## Data

```{r}
#| label: data
incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))
```



* Daily number of attended incidents:\newline 1 October 2015 -- 31 July 2019
* Disaggregated by nature of incidents, priority, the health board managing the service and the control area (or region).
* `r label_comma()(NROW(incident_gthf))` rows observations from `r label_comma()(NROW(attributes(incident_gthf)$key))` time series.

## Data structure

```{r}
#| include: false
data <- data.frame(
  level1 = "Total",
  level2 = c(
    "Central & West", "Central & West", "Central & West",
    "North", "South & East", "South & East", "South & East"
  ),
  level3 = c("HD", "SB", "PO", "BC", "CV", "CT", "AB")
)
# transform it to a edge list!
edges_level1_2 <- data |>
  select(level1, level2) |>
  unique() |>
  rename(from = level1, to = level2)
edges_level2_3 <- data |>
  select(level2, level3) |>
  unique() |>
  rename(from = level2, to = level3)
Cairo::CairoPDF(here::here("figs/group.pdf"), width = 6, height = 6)
rbind(edges_level1_2, edges_level2_3) |>
  igraph::graph_from_data_frame() |>
  ggraph::ggraph(layout = "dendrogram", circular = FALSE) +
  ggraph::geom_edge_diagonal() +
  ggraph::geom_node_point(color = "#dddddd", size = 10) +
  ggraph::geom_node_text(
    aes(label = c(
      "All country",
      "Central & West", "North", "South & East",
      "HD", "SB", "PO", "BC", "CV", "CT", "AB"
    ))
  ) +
  theme_void() +
  theme(
    panel.background = element_rect(fill = "transparent"), # transparent panel bg
    plot.background = element_rect(fill = "transparent"), # transparent plot bg
  )
crop::dev.off.crop()
```

\placefig{0.3}{1.5}{width=7cm}{group.pdf}
\placefig{7.5}{1.5}{width=7.7cm}{group.png}

## Data structure
\fontsize{10}{11}\sf

```{r}
agg_level <- tibble::tribble(
  ~Level, ~`Number of series`,
  "All country", 1,
  "Control", 3,
  "Health board", 7,
  "Priority", 3,
  "Priority * Control", 9,
  "Priority * Health board", 21,
  "Nature of incident", 35,
  "Nature of incident * Control", 105,
  "Nature of incident * Health board", 245,
  "Priority * Nature of incident", 104,
  "Control * Priority * Nature of incident", 306,
  "Control * Health board * Priority * Nature of incident (Bottom level)", 691,
  "Total", 1530
)
knitr::kable(agg_level, booktabs = TRUE, position = "left", linesep = "") |>
  kable_classic(full_width = FALSE)
```

## Aggregated daily incidents

```{r}
incident_gthf |>
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) |>
  autoplot(incident) +
  labs(x = "Date", y = "Incidents")
```

## Daily incidents by control area

```{r}
incident_gthf |>
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) |>
  as_tibble() |>
  select(-nature, -category) |>
  group_by(date, region) |>
  summarise(incident = sum(incident), .groups = "drop") |>
  ggplot(aes(x = date, y = incident, color = factor(region))) +
  geom_line() +
  labs(x="Date", y = "Incidents", color = "Control")
```

## Data incidents by health board

```{r}
incident_gthf |>
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) |>
  as_tibble() |>
  select(-nature, -category) |>
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  labs(y = "Incidents", color = "Health board")
```

## Data incidents by priority

```{r}
incident_gthf |>
  filter(is_aggregated(region) & is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) |>
  mutate(
    category = recode(category, RED = "Red", AMB = "Amber", GRE = "Green"),
    category = factor(category, levels = c("Red", "Amber", "Green"))
  ) |>
  as_tibble() |>
  select(-nature, -region) |>
  ggplot(aes(x = date, y = incident, color = factor(category))) +
  geom_line() +
  scale_color_manual(values = c(Red = "#ff3300", Amber = "#E69f00", Green = "#009e73")) +
  labs(y = "Incidents", color = "Priority")
```

## Data incidents by nature of incident

```{r}
incident_gthf |>
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & !is_aggregated(nature)) |>
  as_tibble() |>
  mutate(nature = as.character(nature)) |>
  group_by(date, nature, lhb) |>
  summarise(incident = sum(incident), .groups = "drop") |>
  ggplot(aes(x = date, y = incident, color = nature)) +
  geom_line() +
  labs(x = "Date", y = "Incidents", color = "Nature of incident")
```

## Data incidents by nature of incident

```{r}
selected_nature <- c("CHESTPAIN", "STROKECVA", "BREATHING", "ABDOMINAL")
incident_gthf |>
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & !is_aggregated(nature)) |>
  as_tibble() |>
  mutate(nature = as.character(nature)) |>
  filter(nature %in% selected_nature) |>
  group_by(date, nature, lhb) |>
  summarise(incident = sum(incident), .groups = "drop") |>
  ggplot(aes(x = date, y = incident, color = nature)) +
  geom_line() +
  labs(x = "Date", y = "Incidents", color = "Nature of incident")
```

## Data features

```{r}
incident_gthf |>
  features(incident, feat_stl) |>
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(alpha = 0.25) +
  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
```

# Forecast methods and evaluation

## Forecasting methods

1. **Naïve**: Empirical distribution of past daily attended incidents.
2. **ETS**: Exponential Smoothing State Space models.
3. **GLM**: Poission Regression with spline trend, day of the week, annual Fourier seasonality, public holidays, school holidays, Christmas Day, New Year's Day.
4. **TSGLM**: Poisson Regression with same covariates plus three autoregressive terms.
5. **Ensemble**: Mixture distribution of 1--4.

## Notation reminder

\begin{textblock}{8.5}(0.2,1.5)
Every collection of time series with linear constraints can be written as
\centerline{\colorbox[RGB]{210,210,210}{$\bY_{t}=\color{blue}\bS\color{red}\bm{b}_{t}$}}
\vspace*{-0.9cm}\begin{itemize}\parskip=0cm\itemsep=0cm
\item $\by_t=$ vector of all series at time $t$
\item $ y_{\text{Total},t}= $ aggregate of all series at time
$t$.
\item $ y_{X,t}= $ value of series $X$ at time $t$.
\item $\color{red}{\bm{b}_t}=$ vector of most disaggregated series at time $t$
\item $\color{blue}{\bS}=$ ``summing matrix'' containing the linear constraints.
\end{itemize}
\end{textblock}

\begin{textblock}{5.7}(11.4,0.1)
\begin{minipage}{4cm}
\begin{block}{}\centering
\begin{tikzpicture}
\tikzstyle{every node}=[ellipse,draw,fill=red!15,inner sep=2pt]
\tikzstyle[level distance=.3cm]
\tikzstyle[sibling distance=12cm]
\tikzstyle{level 1}=[sibling distance=10mm,font=\small,set style={{every node}+=[fill=blue!15]}]
\node{Total}[edge from parent fork down]
 child {node {A}
 }
 child {node {B}
 }
 child {node {C}
 };
\end{tikzpicture}
\end{block}
\end{minipage}
\end{textblock}

\only<1>{\begin{textblock}{5.7}(9.4,2.8)\fontsize{14}{15}\sf
\begin{align*}
\bY_{t}&= \begin{pmatrix}
  y_{\text{Total},t}\\
  y_{A,t}\\
  y_{B,t}\\
  y_{C,t}
  \end{pmatrix}  \\
  &= {\color{blue}\underbrace{\begin{pmatrix}
                1 & 1 & 1 \\
                1 & 0 & 0 \\
                0 & 1 & 0\\
                0 & 0 & 1
                \end{pmatrix}}_{\bS}}
     {\color{red}\underbrace{\begin{pmatrix}
       y_{A,t}\\y_{B,t}\\y_{C,t}
       \end{pmatrix}}_{\bm{b}_{t}}}
\end{align*}
\end{textblock}}
\only<2>{\begin{textblock}{5.7}(9.4,3.2)\fontsize{14}{15}\sf
\begin{alertblock}{}
\begin{itemize}\itemsep=0.1cm
\item Base forecasts: $\hat{\bm{y}}_{T+h|T}$
\item Reconciled forecasts: $\tilde{\bm{y}}_{T+h|T}=\bS\bm{G}\hat{\bm{y}}_{T+h|T}$
\item MinT: $\bG = (\bS'\bm{W}_h^{-1}\bS)^{-1}\bS'\bm{W}_h^{-1}$ where $\bm{W}_h$ is covariance matrix of base forecast errors.
\end{itemize}
\end{alertblock}
\end{textblock}}

## Nonparametric bootstrap reconciliation


## Performance evaluation

* Time series cross-validation: forecast horizon of 1--84 \rlap{days.}
* Forecasts at 43--84 days correspond to planning horizon.

1. Training set: all data to 2018-04-25.
2. Training set: all data to 2018-06-06.

$\hspace*{2cm}\vdots$

10. Training set: all data to 2019-05-09.

## Performance evaluation

\vspace*{0.2cm}\begin{block}{}
\centerline{$\text{MASE} = \text{mean}(|q_{j}|)$}
where
$$
  q_{j} = \frac{e_{j}}
 {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T |y_{t}-y_{t-m}|},
$$
\end{block}

* $e_{j}=$ forecast error for forecast horizon $j$,
* $m = 7$
* $y_t=$ observation for period $t$
* $T=$ size of training set

## Performance evaluation

\vspace*{0.2cm}\begin{block}{}
\centerline{$\text{MSSE} = \text{mean}(q_{j}^2)$}
where
$$
  q^2_{j} = \frac{ e^2_{j}}
 {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
$$
\end{block}

* $e_{j}=$ forecast error for forecast horizon $j$,
* $m = 7$
* $y_t=$ observation for period $t$
* $T=$ size of training set

## Performance evaluation

\vspace*{0.2cm}\begin{block}{}
\centerline{$\text{CRPS} = \text{mean}(p_j)$}
where
$$
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
$$
\end{block}

* $G_j(x)=$ forecasted probability distribution function for forecast horizon $j$
* $F_j(x)=$ true probability distribution function for the same period.



# Results

```{r}
#| include: false
rmsse <- readr::read_rds(here::here("results/rmsse.rds")) |>
  mutate(msse = rmsse^2)
mase <- readr::read_rds(here::here("results/mase.rds"))
crps <- readr::read_rds(here::here("results/crps.rds")) |>
  pivot_wider(names_from = model, values_from = crps) |>
  group_by(method, h, series) |>
  mutate(across(where(is.numeric), ~ .x / naiveecdf)) |>
  ungroup()
```


## Results

```{r}
crps <- crps |>
  pivot_longer(ensemble:tscount, names_to = "model", values_to = "crps")

accuracy <- bind_rows(
  rmsse |> mutate(measure = "msse", accuracy = rmsse^2),
  mase |> mutate(measure = "mase", accuracy = mase),
  crps |> mutate(measure = "crps", accuracy = crps),
) |>
  select(-rmsse, -mase, -crps)

# Plot of average accuracy vs week for each method for Total
acc_summary <- accuracy |>
  filter(
    series %in% c("Total", "Control areas", "Health boards"),
    method == "mint", model != "qcomb"
  ) |>
  mutate(model = case_when(
    model == "naiveecdf" ~ "Naïve",
    model == "ets" ~  "ETS",
    model == "tscount" ~ "TSGLM",
    model == "iglm" ~ "GLM",
    model == "ensemble" ~ "Ensemble"
  )) |>
  mutate(
    measure = factor(measure, levels = c("mase","msse","crps"), labels = c("MASE","MSSE","CRPS")),
    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
    model = factor(model, levels = c("Naïve", "ETS", "TSGLM", "GLM", "Ensemble")),
    week = factor(trunc((h - 1) / 7) + 1)
  ) |>
  group_by(week, model, measure, series) |>
  summarise(accuracy = mean(accuracy), .groups = "drop")


acc_summary |>
  ggplot(aes(x = week, y = accuracy, group = model, col = model)) +
  geom_line() +
  geom_point(size = .5) +
  facet_grid(measure ~ series, scales = "free_y") +
  labs(y = "Average accuracy", x = "Week ahead")
```

## Results
\fontsize{10}{10}\sf

```{r}
# Set minimum in column to bold
set_to_bold <- function(table) {
  for(i in 3:6) {
    best <- which.min(table[[i]])
    table[[i]] <- sprintf(table[[i]], fmt="%1.3f")
    table[[i]][best] <- cell_spec(table[[i]][best], bold=TRUE)
  }
  return(table)
}

msse1 <- rmsse |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble"),
    labels = c("Naïve", "ETS", "GLM", "TSGLM","Ensemble")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT")),
  ) |>
  group_by(method, model, series) |>
  summarise(msse = mean(msse), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = msse) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
kbl(msse1, booktabs = TRUE, escape = FALSE, align="llrrrr") |>
  add_header_above(c(" " = 2, "MSSE" = 4)) |>
  kable_styling(latex_options = c("hold_position"))
```

## Results
\fontsize{10}{10}\sf

```{r}
mase1 <- mase |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42
  ) |>
  mutate(
    model = factor(model,
      levels = c("naiveecdf","ets","iglm","tscount","ensemble"),
      labels = c("Naïve", "ETS", "GLM", "TSGLM","Ensemble")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(mase = mean(mase), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = mase) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
kbl(mase1, booktabs = TRUE, escape = FALSE, align="llrrrr") |>
  add_header_above(c(" " = 2, "MASE" = 4)) |>
  kable_styling(latex_options = c("hold_position"))
```

## Results
\fontsize{10}{10}\sf

```{r}
crps1 <- readr::read_rds(here::here("results/crps.rds")) |>
  filter(method %in% c("mint", "base"),
         model != "qcomb",
         h > 42
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble"),
    labels = c("Naïve", "ETS", "GLM", "TSGLM","Ensemble")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(crps = mean(crps), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = crps) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
kbl(crps1, booktabs = TRUE, escape = FALSE, align="llrrrr") |>
  add_header_above(c(" " = 2, "CRPS" = 4)) |>
  kable_styling(latex_options = c("hold_position"))
```

## Conclusions
